\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-28 paper/algorithm.tex
paper: rewrote dominance definition, changed plan cache explanation to partial order

\section{Algorithm}
In order to correctly plan around missing values, we first extend the relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
aggregations. The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
quality of the imputations performed and the runtime performance. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). Our imputation planning algorithm is
agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
that while our algorithm is exponential in the number of joins(\Cref{sec:complexity}), in practice, planning times are not a concern.

@@

\begin{definition}
We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$ if $(l \leq l' \land t < t') \lor (l < l' \land t \leq t')$. We denote this as $\doms{p}{p'}$. 
\end{definition}

@@

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
                                            \text{ where } \\
                                           P' &= \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
                                                                     \forall q' \in \llbracket Q[T] \rrbracket, \textsc{Dirty}(q')=\textsc{Dirty}(q).  \doms{q} {q'} 
                                                                   \end{aligned}
                                           \right\}
  \end{align*}
  \caption{Semantics of a plan cache $Q$}
  \label{fig:semantics-plan-cache}
\end{figure}

\todobox{jose rewrite this clearly}{
While $\textsc{Loss}$ is not monotonically increasing in our model, we maintain a collection of sound and complete Pareto frontiers by 
computing these specific to each dirty set. That is to say, that the combination of $\textsc{Loss}(p) \times \textsc{Dirty}(p)$ for some
plan $p$ behaves monotonically, as the dirty-set is strictly decreasing through the application of $\delta, \mu$ operations. 
The final Pareto frontier corresponds to the collection of plans associated with the empty dirty set.}

>>>>

\section{Algorithm}
In order to correctly plan around missing values, we first extend the relational algebra operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate $g$) 
with two new operators (\Cref{sec:operators}). We then define the search space (\Cref{sec:search-space}) of plans to encompass non-trivial operations, such as joins and 
aggregations. Soundness of query execution is provided by our main design invariant, which guarantees traditional relational
algebra operators must never observe a missing value in any attribute that they operate on directly (\Cref{sec:placement}). 
The decision to place imputation operators is driven by our cost model (\Cref{sec:cost-model}), which characterizes a plan based on estimates for the 
quality of the imputations performed and the runtime performance. Our imputation planning algorithm is
agnostic to the type of imputation used (\Cref{sec:imputation}). Finally, we show
that while our algorithm is exponential in the number of joins(\Cref{sec:complexity}), in practice, planning times are not a concern.

@@

\begin{definition}
We say that a plan $p$ with cost $\langle l, t \rangle$ dominates a plan $p'$ with cost $\langle l', t' \rangle$ if $\textsc{Dirty}(p) = \textsc{Dirty}(p') \land ((l \leq l' \land t < t') \lor (l < l' \land t \leq t'))$. We denote this as $\doms{p}{p'}$. 
\end{definition}

@@

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned} Q \cup \{(T,\ P') \} \end{aligned}\\
                                            \text{ where } \\
                                           P' &= \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p),\\
                                                                     \forall q' \in \llbracket Q[T] \rrbracket.  \doms{q} {q'} 
                                                                   \end{aligned}
                                           \right\}
  \end{align*}
  \caption{Semantics of a plan cache $Q$}
  \label{fig:semantics-plan-cache}
\end{figure}

The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Loss}, \textsc{Time} \rangle$ to
collect sound and complete Pareto frontiers. For plans with the same dirty set, we can treat $\langle \textsc{Loss}, \textsc{Time} \rangle$
as a total order, allowing us to appropriately select the frontier. Meanwhile, plans with different $\textsc{Dirty}$ sets cannot be compared. The final
Pareto frontier corresponds to the collection of plans associate with the empty dirty set.
